---
title: "Least Squares By Springs"
author: "Michael Friendly"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Least Squares By Springs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r nomessages, echo = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.height = 5,
  fig.width = 5
)
options(digits=4)
par(mar=c(3,3,1,1)+.1)
```

```{r setup, include=FALSE}
library(ggsprings)
library(tibble)
library(tinytable)
library(dplyr)
```

## Introduction

- Hooke's Law, potential energy and force

- Functions, energy, force

```{r echo = FALSE}
tbl <- tibble::tribble(
  ~Mathematics,   ~Physical,
  "function, $f (x)$",          "potential energy, $P(x)$",
  "derivative, $f\\prime (x)$",  "force, $F(x) = -P\\prime(x)$",
  "$\\argmin_x f(x) \\implies f\\prime (x) = 0$",
          "equilibrium, $\\argmin_x P(x) \\implies F (x) = 0$"
  )

tt(tbl)
```


## Mean of a sample

- History of sample mean ...


The sample mean has several nice physical analogs, which stem from the properties that 

* the sum of deviations ($e_i$) of observations ($x_i$) from the mean $\bar{x}$ equals 0: $\Sigma_i e_i =\Sigma_i ( x_i - \bar{x} ) = 0$.

* the sum of squared deviations is the smallest for any choice of $\bar{x}$, $\bar{x} = \argmin_{\bar{x}} ( x_i - \bar{x} )^2$

### Demonstration using springs


Create a set of observations ...
```{r mean1}
set.seed(1234)
N <- 8
df <- tibble(
  x = runif(N, 1, 10),
  y = seq(1, N)
)

means <- colMeans(df) 
xbar <- means[1] |> print()
```

First, set the tension of the spring to be the squared deviation of the observation from the mean.
```{r mean2}
df <- df |>
  mutate(tension = (x - xbar)^2,
         diameter = 0.2)
```

Then, visualize with springs: 

```{r mean-springs}
ggplot(df, aes(x=x, y=y)) +
  geom_point(size = 5, color = "red") +
  geom_segment(x = xbar, xend = xbar,
               y = 1/2,  yend = N + 1/2,
               linewidth = 3) +
  geom_spring(aes(x = x, xend = xbar,
                  y = y, yend = y,
                  tension = tension,
                  diameter = diameter),
              color = "blue",
              linewidth = 1.2) +
  labs(x = "Value (x)",
       y = "Observation number") +
  ylim(0, N+2) +
  scale_y_continuous(breaks = 1:N) +
  annotate("text", x = xbar, y = N + 1,
           label = "Movable\nrod", size = 5,
           , lineheight = 3/4) +
  theme_minimal(base_size = 15)
```

## Bivariate centroid

For a bivariate sample, $(x_i, y_i)$, the centroid point, $(\bar{x}, \bar{y})$ is realized physically as the result of attaching
springs between the fixed points and a free, movable point. Each point has potential energy proportional to its squared distance
from the movable one, where the forces balance (sum to zero).

Set this up for a sample of 10 points, uniformly distributed on (1, 10).
```{r cent1}
set.seed(1234)
N <- 10
df <- tibble(
  x = runif(N, 1, 10),
  y = runif(N, 1, 10)
)

means <- colMeans(df)
xbar <- means[1]; ybar <- means[2]
```

Set the tension:

```{r cent2}
df <- df |>
  mutate(tension = (x - xbar)^2 + (y - ybar)^2,
         diameter = 0.4)
```

Visualize the springs:

```{r centroid-springs}
ggplot(df, aes(x=x, y=y)) +
  geom_point(size = 5, color = "red") +
  geom_spring(aes(x = x, xend = xbar,
                  y = y, yend = ybar,
                  tension = tension / 5,
                  diameter = diameter),
              color = "blue",
              linewidth = 1.2) +
  geom_point(x = xbar, y = ybar, 
             size = 7,
             shape = 15,
             color = "black") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(breaks = 1:10) +
  theme_minimal(base_size = 15) 
```

### Animate this?

One way to animate this would be to imagine the springs acting one at a time, sequentially on the position of the moving point, and looping until nothing changes. Not sure how to program this with `gganimate`.


## Least squares regression

